Let's talk about Thurstone & Co.: An information-theoretical model for comparative judgments, and its statistical translation

### 2024-10-22 index.qmd
section:issues; process:write

---
title: "Causes and effects in Dichotomous Comparative Judgments: an information-theoretical system of plausible mechanism"
author:
  - name: 
      given: Jose Manuel
      family: Rivera Espejo
    orcid: 0000-0002-3088-2783
    url: https://www.uantwerpen.be/en/staff/jose-manuel-rivera-espejo_23166/
    email: JoseManuel.RiveraEspejo@uantwerpen.be
    corresponding: true
    affiliation:
      - name: University of Antwerp
        department: Training and education sciences
        group: Edubron
  - name: 
      given: Tine
      family: van Daal
      non-dropping-particle: van
    orcid: https://orcid.org/0000-0001-9398-9775
    url: https://www.uantwerpen.be/en/staff/tine-vandaal/
    email: tine.vandaal@uantwerpen.be
    corresponding: false
    affiliation:
      - name: University of Antwerp
        department: Training and education sciences
        group: Edubron
  - name: 
      given: Sven
      family: De Maeyer
      non-dropping-particle: De
    orcid: 0000-0003-2888-1631
    url: https://www.uantwerpen.be/en/staff/sven-demaeyer/
    email: sven.demaeyer@uantwerpen.be
    corresponding: false
    affiliation:
      - name: University of Antwerp
        department: Training and education sciences
        group: Edubron
  - name: 
      given: Steven
      family: Gillis
    orcid: 
    url: https://www.uantwerpen.be/nl/personeel/steven-gillis/
    email: steven.gillis@uantwerpen.be
    corresponding: false
    affiliation:
      - name: University of Antwerp
        department: Linguistics
        group: Centre for computational linguistics, psycholinguistics, and sociolinguistics (CLiPS)
funding: 
  statement: "The project was founded through the Research Fund of the University of Antwerp (BOF)."
keywords:
  - causal inference
  - probability
  - Thurstone
  - comparative judgement
  - directed acyclic graph
  - structural causal models
  - statistical modeling
abstract: |
  (to do)
key-points:
  - (to do)
date: last-modified
bibliography: references.bib
notebook-links: global
lightbox: true
citation: true
  # type: article-journal
  # container-title: "Psychometrika"
  # doi: ""
  # # url: https://example.com/summarizing-output
execute: 
  cache: true
  # eval: true
  echo: false
  # output: true
  # include: true
  warning: false
  error: false
  message: false
---

@@

<!-- recording -->
<!-- Part 1; 00:00:00 - 00:09:10 -->
<!-- Part 2; 00:14:30 - 00:21:00 -->

<!-- 1. Where are we now with DCJ? -->

<!-- Notes: -->
<!-- - In the past decade there has been a growing number of studies that document and analyze the merits of DCJ as an assessment tool. Define a bit DCJ. -->
<!-- - It has been a growing field, where people has tested some important aspects of the method. In general, is an iterative process. People are iterating on particular aspects of the method, such as: -->
<!-- a. issues within the method as such, critical analyze some aspect of the method -->
<!-- b. practical use of the method, generate scores with the BTL, and use this scores with some hypotheses, or descriptions. -->

Over the past decade, numerous studies have documented the effectiveness of the *comparative judgment* (CJ) method [@Thurstone_1927] for assessing competencies and traits. These studies have evaluated CJ from two main perspectives: its ability to produce reliable and valid trait scores, and its practical applicability. In terms of reliability and validity, research has shown that CJ can generate precise and consistent scores that accurately represent the traits being measured. Notable contributions in this research area include studies by Pollit [-@Pollitt_2012a; -@Pollitt_2012b], @Whitehouse_2012, @vanDaal_et_al_2016, @Lesterhuis_2018, @Bramley_et_al_2019, @Verhavert_et_al_2019, @Crompvoets_et_al_2022, and @Bouwer_et_al_2023. Regarding practical applicability, several studies have highlighted the method's versatility in both educational and non-educational contexts, presenting it as an efficient and effective alternative for measurement and evaluation. Key examples in this research area include the works of @Jones_2015, @Bartholomew_et_al_2018, @Jones_et_al_2019, @Marshall_et_al_2020, @Bartholomew_et_al_2020, and @Boonen_et_al_2020.

<!-- 2. But what are the problems of DCJ as a method? -->

<!-- Notes:  -->
<!-- - But what it is the problem with this way of using DCJ (just one paragraph) -->
<!-- a. measurement model is based on case V -->
<!--   a.1 It does not consider the effects of judges (different perceptions per stimuli) -->
<!--   a.2 It does not consider that stimuli may be nested in individuals -->
<!-- b. in all literature there is a disconnection between measurement model and the structural part (common practical use) -->
<!-- c. what about the experimental design?, what are the implications of certain comparison algorithms? -->

Despite the growing number of CJ studies, the unsystematic and fragmented research approaches employed in the literature have overlooked several critical issues concerning CJ. These issues can be categorized into concerns about the method's structural model, measurement model, and experimental design.

A key issue in the first category is the apparent disconnect between the method's structural and measurement models. A common practice in CJ literature involves performing data analysis and hypothesis testing on scores previously estimated using the Bradley-Terry-Luce (BTL) model [@Bradley_et_al_1952; @Luce_1959]. These studies use the scores or their transformations to identify 'misfit' judges and stimuli [@Pollitt_2012b; @vanDaal_et_al_2017], detect 'bias' in judges' ratings [@Pollitt_et_al_2003; @Pollitt_2012b], or test various hypothesis [@Bramley_et_al_2019; @Boonen_et_al_2020; @Bouwer_et_al_2023; @vanDaal_et_al_2017; @Jones_et_al_2019; @Gijsen_et_al_2021]. However, since CJ scores are parameter estimates with inherent uncertainty, the statistical literature suggests that separating the analysis from this uncertainty may artificially inflate the precision and power of the results. This, in turn, could increase the risk of committing a type I error, where a null hypothesis is wrongly rejected [@McElreath_2020].

# Theory {#sec-theory}
<!-- 21:00 recording -->

## Let's talk about Thurstone co. {#sec-theory-thurstone}

<!-- Notes: -->
<!-- - But what it is the problem with this way of using DCJ (just one paragraph) -->
<!-- a. measurement model is based on case V -->
<!--   a.1 It does not consider the effects of judges (different perceptions per stimuli) -->
<!--   a.2 It does not consider that stimuli may be nested in individuals -->
<!-- b. in all literature there is a disconnection between measurement model and the structural part (common practical use) -->
<!-- c. what about the experimental design?, what are the implications of certain comparison algorithms? -->

<!-- Use the SEM description of measurement model and structural model. -->

<!-- The issues fall into three categories: those related to the measurement model, the structural model, and the experimental design. -->

@@

<!-- (a) related to Case 3, it assumes the correlation between stimuli is zero, this translates into the cancellation of judges effects by mean of opposing and equally weigthed 'mood' and 'simultaneous contrasts' effects. This is demonstrated using the additivive nature of the logit scale, which helps to cancel 'bias' judges effects. -->
<!-- This is the case of confounding, and @Andrich_1978 already showed that if judges effects are parametrized, it gets eliminated experimentally in the paired comparison method.  -->

@@

<!-- Structural model -->

<!-- Regarding this, it is apparent that in all literature there is a disconnection between measurement model and the structural part, where hypothesis and summaries are performed. -->

<!-- The current literature estimates the CJ scores to later used them into other analysis, such as investigating bias and misfits among judges, or for the comparison of groups of individuals. This ignores measurement error. -->

@@

<!-- present the evidence of bias effects of adaptive algorithms, the not clear view of how many comparisons per (sub)units are required. Adhoc rules of thumb. -->

## A scientific model for the CJ {#sec-theory-scientific}

@@

@@

>>>>

---
title: "Let's talk about Thurstone & Co.: An information-theoretical model for comparative judgments, and its statistical translation"
author:
  - name: 
      given: Jose Manuel
      family: Rivera Espejo
    orcid: 0000-0002-3088-2783
    url: https://www.uantwerpen.be/en/staff/jose-manuel-rivera-espejo_23166/
    email: JoseManuel.RiveraEspejo@uantwerpen.be
    corresponding: true
    affiliation:
      - name: University of Antwerp
        department: Training and education sciences
        group: Edubron
  - name: 
      given: Tine
      family: van Daal
      non-dropping-particle: van
    orcid: https://orcid.org/0000-0001-9398-9775
    url: https://www.uantwerpen.be/en/staff/tine-vandaal/
    email: tine.vandaal@uantwerpen.be
    corresponding: false
    affiliation:
      - name: University of Antwerp
        department: Training and education sciences
        group: Edubron
  - name: 
      given: Sven
      family: De Maeyer
      non-dropping-particle: De
    orcid: 0000-0003-2888-1631
    url: https://www.uantwerpen.be/en/staff/sven-demaeyer/
    email: sven.demaeyer@uantwerpen.be
    corresponding: false
    affiliation:
      - name: University of Antwerp
        department: Training and education sciences
        group: Edubron
  - name: 
      given: Steven
      family: Gillis
    orcid: 
    url: https://www.uantwerpen.be/nl/personeel/steven-gillis/
    email: steven.gillis@uantwerpen.be
    corresponding: false
    affiliation:
      - name: University of Antwerp
        department: Linguistics
        group: Centre for computational linguistics, psycholinguistics, and sociolinguistics (CLiPS)
funding: 
  statement: "The project was founded through the Research Fund of the University of Antwerp (BOF)."
keywords:
  - causal inference
  - probability
  - Thurstone
  - comparative judgement
  - directed acyclic graph
  - structural causal models
  - statistical modeling
abstract: |
  (to do)
key-points:
  - (to do)
date: last-modified
bibliography: references.bib
notebook-links: global
lightbox: true
citation: true
  # type: article-journal
  # container-title: "Psychometrika"
  # doi: ""
  # # url: https://example.com/summarizing-output
execute: 
  cache: true
  # eval: true
  echo: false
  # output: true
  # include: true
  warning: false
  error: false
  message: false
---

@@

<!-- recording 1 -->
<!-- Part 1; 00:00:00 - 00:09:10 -->

<!-- recording 2 -->
<!-- Part 2; 00:00:00 - 00:17:30 -->

<!-- 1. Where are we now with DCJ? -->
Over the past decade, numerous studies have documented the effectiveness of the *comparative judgment* (CJ) method [@Thurstone_1927] for assessing competencies and traits. These studies have evaluated CJ from three main perspectives: its ability to produce reliable and valid trait scores, its practical applicability, and its time efficiency. Research on reliability and validity shows that CJ can generate precise and consistent scores [@Pollitt_2012a; @Pollitt_2012b; @Coertjens_et_al_2017; @Goossens_et_al_2018; @Verhavert_et_al_2019; @Crompvoets_et_al_2022; @Bouwer_et_al_2023] that accurately represent the traits being measured [@Whitehouse_2012; @vanDaal_et_al_2016; @Lesterhuis_2018; @Bouwer_et_al_2023]. Research on practical applicability highlights CJ's versatility across both educational and non-educational contexts, presenting it as an efficient and effective alternative for measurement and evaluation [@Pollitt_2004; @Jones_2015; @Bartholomew_et_al_2018; @Jones_et_al_2019; @Marshall_et_al_2020; @Bartholomew_et_al_2020; @Boonen_et_al_2020]. Lastly, research on time efficiency suggests that CJ can offer at least equal, if not significant, time savings when evaluating stimuli compared to traditional marking methods [@Pollitt_2012a; @Pollitt_2012b; @Coertjens_et_al_2017; @Goossens_et_al_2018].

<!-- 2. But what are the problems of DCJ as a method? -->
Nevertheless, despite the growing number of CJ studies, the unsystematic and fragmented research approaches employed in the literature have overlooked several critical issues concerning the method. These issues fall into three main categories: the disconnect between CJ's structural and measurement model, the over-reliance on Thurstone's Case 5 [-@Thurstone_1927] in its measurement model, and the role of the comparison algorithms and the number of comparisons on the method's reliability and validity. In the following sections, each issue will be discussed in detail, followed by the introduction of a theoretical model and its translation into a statistical model that addresses all three concerns simultaneously.

# Overlooked issues in CJ literature {#sec-theory-issues}

## The disconnect between structural and measurement model

<!-- 1. The BTL model is CJ's measurement model -->
In the CJ literature, the Bradley-Terry-Luce (BTL) model [@Bradley_et_al_1952; @Luce_1959] serves as the measurement model for CJ because it specifies how latent variables are estimated from observed ones [@Hoyle_et_al_2023; @Kline_et_al_2023]. In a CJ study, multiple judges conduct several rounds of pairwise comparisons to evaluate the relative manifestation of a trait between a pair of stimuli. This evaluation results in a dichotomous outcome that indicates which stimulus is perceived to exhibit a higher degree of the trait. The BTL model then uses these observed outcomes to estimate scores that represent the latent trait of interest [@Pollitt_2012a; @Pollitt_2012b; @Whitehouse_2012; @Jones_2015; @vanDaal_et_al_2016; @Lesterhuis_2018; @Boonen_et_al_2020; @Bouwer_et_al_2023].

<!-- 2. But there is no structural model, only post estimation analyses -->
Moreover, in the CJ literature, it is common practice for BTL-generated scores, or their transformations, to undergo additional data analysis and hypothesis testing. Researchers have utilized these scores in independent analyses to identify 'misfit' judges and stimuli [@Pollitt_2012b; @vanDaal_et_al_2017; @Goossens_et_al_2018], detect biases in judges' ratings [@Pollitt_et_al_2003; @Pollitt_2012b], calculate correlations with other scoring methods [@Goossens_et_al_2018; @Bouwer_et_al_2023], or test hypotheses related to the trait of interest [@Bramley_et_al_2019; @Boonen_et_al_2020; @Bouwer_et_al_2023; @vanDaal_et_al_2017; @Jones_et_al_2019; @Gijsen_et_al_2021].

<!-- 3. But what is the problem with this? -->
<!-- However, the statistical literature cautions when using estimated scores for further independent data analysis and hypothesis testing. Since BTL-generated scores are parameter estimates that carry inherent uncertainty, separating the subsequent analyses from this uncertainty may lead to an artificial inflation of the precision and power of the analyses and tests. Consequently, this practice could increase the risk of committing a type I error, wherein a null hypothesis is incorrectly rejected [@McElreath_2020]. An appropriate approach to address this issue, consistent with the principles of Structural Equation Modeling (SEM) [@Hoyle_et_al_2023; @Kline_et_al_2023], is that these analyses should be conducted at the structural level rather than as independent analyses. Consequently, an approach that integrates the structural and measurement model should be followed. -->

<!-- However, there is often a lack of integration between the structural model, which captures the relationships among the variables, and the measurement model, which generates the CJ scores. This disconnect can lead to inaccurate inferences about the traits being measured. -->

## Case 5 and the measurement model

<!-- a. measurement model is based on case V -->
<!--   a.1 It does not consider the effects of judges (different perceptions per stimuli) -->
<!--   a.2 It does not consider that stimuli may be nested in individuals -->

<!-- (a) related to Case 3, it assumes the correlation between stimuli is zero, this translates into the cancellation of judges effects by mean of opposing and equally weigthed 'mood' and 'simultaneous contrasts' effects. This is demonstrated using the additivive nature of the logit scale, which helps to cancel 'bias' judges effects. -->
<!-- This is the case of confounding, and @Andrich_1978; @Wainer_et_al_1978 already showed that if judges effects are parametrized, it gets eliminated experimentally in the paired comparison method.  -->

<!-- @Pollitt_et_al_2003 In Thurstone’s method each judge’s standard cancels out when a comparison is made, and any misfit between data and model must indicate a difference of some other kind. This is studied by analysing the residuals that remain after the model is fitted to the data. Whenever one script is judged to have a higher value than a competing one it is expected to ‘win’ the comparison: if it does the residual will be small, but if it loses then the residual for that judge’s comparison of those two scripts will be large. Any patterns of large residuals will indicate some sort of bias – in this context bias simply means a source of systematic, rather than random, variability. indicates that one judge – Judge 5 – is significantly biased in some way. (Consider the problem of using adhoc methods) -->

@@

@@

<!-- All studies use SSR as a measure of reliability -->

## The efficiency and reliability of comparison algorithms

@@

<!-- @Bramley_et_al_2019 and @Bramley_2015 says adaptivity inflates reliability. present the evidence of bias effects of adaptive algorithms, the not clear view of how many comparisons per (sub)units are required. Adhoc rules of thumb. -->

<!-- @Gray_et_al_2024 entropy-driven active learning pair-selection algorithm to compare stimuli, vs random and no repeating pairs -->

<!-- @DeVrindt_et_al_2024 cold start problem to optimize time efficiency. -->

<!-- @Mikhailiuk_et_al_2021 proposed a fully Bayesian active sampling strategy for pairwise comparisons via Approximate Message Passing and Information Gain Maximization. ASAP computes the full posterior distribution, which is crucial to achieving accurate EIG estimates, and thus the accuracy of active sampling. We recommend ASAP, as it offered the highest accuracy of inferred scores compared to existing methods in experiments with real and synthetic data. -->

<!-- @Verhavert_et_al_2022 The current article proposes a new adaptive selection algorithm using a previously calibrated reference set. Using a reference set should eliminate the reliability inflation. It is proven that this adaptive selection algorithm is more efficient without reducing the accuracy of the results and without increasing the standard deviation of the assessment results. Akin to BIB designs and more specifically to BTIB designs  -->

<!-- @Coertjens_et_al_2017 assess reliability and time investment. We conclude that pairwise comparison requires a similar time investment from evaluators as the criterion list method. We want to note that the outcomes in this study depend heavily on the design. First, the time taken for an assessment is strongly determined by the length of the criteria list used and the length of the texts (Breland, 1983). Not adaptive -->

<!-- Although the process of pairwise comparisons is quite easy and fast for raters to apply (@Laming_2004; @Pollitt_2004), even without training, the increase in time investment that is needed for CJ decreases its feasibility in educational practice. -->

<!-- Messick_1989 talks about efficiency and feasibility for application  -->

<!-- @Bramley_2008 proposes ranking instead of comparisons -->

<!-- talk about the number of comparisons @Verhavert_et_al_2019, @Crompvoets_et_al_2022, @Pollitt_2004 -->

<!-- @Pollitt_2012b concludes that speed is not related to the quality of assessments in pairwise comparisons -->

## What is the solution

# Theory {#sec-theory}
<!-- recording 1 -->
<!-- Part 2; 00:14:30 - 00:21:00 -->

## A theoretical model for the CJ {#sec-theory-scientific}

@@

<!-- @Pritikin_2020 and @Gray_et_al_2024 bayesian modeling attempts -->

@@

<!-- Pollitt (2012a) argues that assessors in pairwise comparisons need less training, given intuitive comparison and decision making. However, there is a distinct lack of research on the role of training.  -->
